\documentclass[a4paper, 11pt]{article}

\usepackage{amsmath}
\usepackage{tikz}
\usetikzlibrary{intersections}

\begin{document}

\section{Root move acceptance probability}

The root move acts on a state vector $x=\{T;n_A,t_A,c_A;n_B,t_B,c_B\}$
where $T$ is the age of the root, $n_A$ and $n_B$ are the numbers of
migration events along each of the two root-connected edges, the
vectors $t_A$ and $t_B$ contain the times of these events and the
vectors $c_A$ and $c_B$ similarly contain the destination colours of
the events.

The subtree slide move draws a new state $x'=\{T',t_A,t'_B\}$
from the following proposal density:
\begin{align}
  q(x'|x) &= P(T',\vec{t}'_A,n'_A,\vec{t}'_B,n'_B|x) \nonumber \\
& = P(T'|T)P(n'_A|T',T)P(n'_B|T',T)P(\vec{t}_A|n'_A)P(\vec{t}_B|n'_B)
\end{align}

\subsection{Tree height proposal}

The new tree height is selected by drawing $f'\sim g_\gamma=U(1/\gamma,\gamma)$
then setting $T'=fT$. We can then write
\begin{equation}
  P(T'|T)=\int df'\delta(T'-f'T)g_\gamma(f')
\end{equation}

\subsection{Migration count proposal}

The new migration counts $n'_A$ and $n'_B$ are chosen from Poissonian
distributions with means of $T'\mu$, where $\mu$ is a tuning
parameter.  That is,
\begin{equation}
  P(n'_A|T',T)=e^{-T'\mu}\frac{(T'\mu)^{n'_A}}{n'_A!}
\end{equation}
and similarly for $P(n'_B|T',T)$.

\subsection{Migration times proposal}

Each of the new migration times in $t'_A$ is chosen by drawing $n'_A$
values $\tau'_A$ independently from $U(0,T')$, then sorting them from smallest
to largest.  If we denote the sorting function $t'_A=S(\tau'_A)$ we
can write
\begin{align}
  P(t'_A|n'_A) &= \int_{[0,T']^{n'_A}}
  d^{n'_A}\tau'_A\delta(t'_A-S(\tau'_A))(T')^{-n'_A}\nonumber\\
&= \frac{n'_A!}{(T')^{n'_A}}
\end{align}
The proposal $P(t'_B|n'_B)$ is defined similarly.

\subsection{Migration colours proposal}

Each migration event results in a change of colour at that point in
the lineage. The new colour is selected at random from those remaining
after the present colour is excluded. That is
\begin{equation}
  P(c'_A|n'_A) = \frac{1}{(N-1)^{n'_A}}
\end{equation}
where $N$ is the total number of available colours. $P(c'_A|n'_B)$ is
defined similarly.



\end{document}

Reviewed-by: Tim Vaughan <tgvaughan@gmail.com>
Tested-by: Tim Vaughan <tgvaughan@gmail.com>
